### 2015-12-05 paper/4-performance.tex
Update lost profiles. Expand 4/6 a bit more + cite GVT futures.

%TODO say that you will now prove your claims about the aproach having nearly no impact
%TODO and potentially even claim that you are faster than adevs for some situations, due to the algorithms from PythonPDEVS and further features
%TODO briefly give some hardware info on the machine (CPU, OS, RAM), used software (version of adevs, GCC, ...), and about number of samples/variance
\subsection{Sequential Simulation}
%TODO we start with sequential simulation results, because this is still base case
%TODO compare CPU time with adevs for some benchmarks
%TODO and explain why there is such a difference
%TODO (spend some time here, as this is also important)
\subsection{Preamble}
The benchmark models are identical in functionality (between adevs/dxex versions). Dxex uses integer timestamps, adevs uses floating point. Tests show there is little difference performance wise (both types fit in a machine word)\\
Benchmarks ran on a i7-2670QM, 16GiB Ram, Arch Linux (4.2.5 kernel). Analysis of the benchmark results is done using call-graphs generated by perf, a selection of those call-graphs can be found in the code repository.
\subsubsection{Devstone}
The Devstone \cite{DEVStone} benchmark is highly hierarchical, using Direct connection the dxex kernels can exploit this, whereas adevs needs to walk the structure of the model to pass events.\\
\begin{figure}[h]
	\includegraphics[width=.5\textwidth]{fig/fig1.eps}
	\label{fig1.eps}
\end{figure}

\subsubsection{PHold}
PHold \cite{PHOLD} is a parallel oriented benchmark, sequential runtime is measured only as a baseline. The usage of random nr generators takes up a significant amount of runtime in this model.

\subsubsection{Interconnect}
Interconnect \cite{van2013research} is a benchmark where all models broadcast, creating a complete graph in terms of dependencies between models. As the model count increases, we see the expected quadratic increase in runtime in both kernels, but an increasing penalty for dxex w.r.t. adevs. Profiling shows this is entirely to the heap allocation of messages by dxex, even though minimized by using memory pools, it remains significant.\\
\begin{figure}[h]
	\includegraphics[width=.5\textwidth]{fig/fig3.eps}
	\label{fig3.eps}
\end{figure}

\subsection{Parallel Simulation}
%TODO explain that we will now compare different synchronization protocols
%TODO compare different synchronization protocols with adevs' conservative, showing that we are faster (or at least, not that much slower) than adevs
% split: dstone/pqueue are dag's
\subsubsection{Devstone}
% Initial.
The flattened models are allocated to kernels by giving each kernel a distinct section of the chain, resulting in a low ratio of inter-kernel to intra-kernel messages. For optimistic, this can cause more reverts since the kernels will start to drift faster as the modelcount increases. Furthermore, optimistic is quite sensitive to an increase in kernels, since the delay before a revert propagates increases. Of note as well is the warm-up time this benchmark requires, for $n=d\times w$ models, it takes $\textup{timeadvance}()*n$ transitions to activate the last model in the chain. For parallel this can be reduced to $\frac{ta()*n}{kernels}$ before the last kernel becomes active.\\
\begin{figure}[ltbh]
	\includegraphics[width=.5\textwidth]{fig/fig2.eps}
	\label{fig2.eps}
\end{figure}
\subsubsection{PHold}
%Phold
In Phold allocation is specified in the benchmark itself. Each kernel manages a single node with a constant set of sub-nodes. The parameter R determines the percentage of remote destination models.\\
The dynamic dependency graph is a very sparse version of the static dependency graph, penalizing conservative. Lookahead is $\epsilon$, so conservative spends most of its time crawling in steps of $\epsilon$. Since the dependency graph between kernels is a complete graph, this is not a simulation that scales in our implementation. For N kernels, each kernel has to query the null-time of N-1 kernels, resulting in O($N^2$) polling behaviour. In a non-cyclic simulation with a non-trivial lookahead (e.g. devstone), that choice however does pay off. Adevs' lesser performance is due in part to their lookahead management, which after profiling shows to spend a non-trivial amount of time in exception handling code. 
Optimistic suffers little from the above problems, however due to the high interconnectivity a cascading revert is still possible. With the percentage of remotes (R) $~$ 100 phold reflects interconnect in behaviour, which is why the R value is not dimensioned here. A revert is very expensive in PHold due to our usage of C++11's random nr generators. The cost of a revert is dominated by the recalculation of destination models, not in allocating/deallocating states/messages. Again, this could be significantly reduced using lazy-cancellation. Once a revert happens the drift between the kernels increases fast, increasing the likelihood of more reverts. Despite all this, optimistic can quickly exploit uncertainty that slows down conservative in this benchmark for low R values.
\subsubsection{Interconnect}
In Interconnect
the set of atomic models form a complete graph (w.r.t. connections), each model broadcasts messages to the entire set. \\
Allocation cannot avoid cycles between the kernels, the resulting dependency graph between kernels remains a complete graph. The runtime dependency graph is almost immediately equal to the static dependency graph.
Conservative still faces the same issues as in PHold, with the key difference that for a fixed time advance lookahead is equal to the timespan between transitions. The scaling issue is identical as in PHold.\\
In this benchmark optimistic runs very quickly out of memory. With c the kernels N atomic models, a single revert undoing k transitions will lead to $(N-1)\times k$ messages that need to be recreated, plus $(N-1) \frac{c-1}{c} \times k$ anti-messages that need to be send.

\subsubsection{Priority network model}
%TODO show a single plot which varies a parameter that changes the ideal synchronization protocol
The priority benchmark is composed of a single server generating a stream of $0<=m<=n$ messages at fixed time intervals, interleaved with a probability p for a priority message, to n receivers. \\ This defaults lookahead for the receivers to $\epsilon$ but this time there is no scaling effect, nor are there cycles in the dependency graph. This model therefore highlights the basic strengths/weaknesses of both synchronization protocols. Receiving models are allocated on another kernel than the server, and have a internal transition so that they will not wait for the incoming messages. 
\input pqueue.tex \\
A key difference here with the other benchmarks is that a state (in the Receiver instances) is very cheap to copy/create. The kernel holding the server will never revert since it is a source in the dependency graph. Optimistic will therefore not suffer the same performance hit in recreating states as it does in PHold. The overhead in optimistic is entirely due to the factor m, which will quickly dominate in increasing buffers of received messages.
\begin{figure}[ltbh]
	\includegraphics[width=.5\textwidth]{fig/fig5.eps}
	\label{fig5.eps}
\end{figure}

%TODO (keep explaining as much as space permits, as this is the evaluation of the core contribution)
\subsection{Memory Usage}
%TODO compare memory usage with adevs for some benchmarks
\subsubsection{Platform and tools}
Both dxex and adevs use tcmalloc as memory allocator. Additionally, dxex uses memory pools to further reduce the frequency of expensive system calls (malloc/free/sbrk/mmap/...). Tcmalloc will only gradually release memory back to the OS, whereas our pools will not do so at all. If memory has been allocated once, it is from a performance point of view better to keep that memory in the pool. This is one reason why memory utilization is best measured by peak allocation. Profiling is done using Valgrind's massif tool. \cite{Nethercote:2007:VFH:1273442.1250746}
Platform used has an i5-3317U Intel cpu and 8GiB RAM with a page size of 4,096KiB, running Fedora 22 (kernel 4.2.6).\\
\subsubsection{Measure}
Adevs passes messages by value (in container by reference), we pass by pointer. The runtime effects of this choice are already demonstrated in the interconnect benchmark, so in this section, we measure memory usage in number of allocated pages combinin text, stack and heap memory for the program profiled. For the OS and/or user this is the actual memory footprint of the application. It is important to note that, especially in the case of optimistic, not all this memory is always in use by the kernels. During simulation, the pools will generally not return deallocated memory to the OS, but keep it for later reuse.
\subsubsection{Results}
\paragraph*{Devstone}
\begin{table}[lhtb]
	\centering
	\caption{Devstone 40x40 t5e5, unit MiB, 4 kernels (if parallel)}
	\label{dtone_mem}
	\begin{tabular}{| l | l | l | l | l |}
		\hline
		Adevs & AdevsCon &DX &DXCon &DXOpt\\ \hline
		44 & 70 & 42 & 75 & 363  \\ \hline
	\end{tabular}
\end{table}
Since conservative passes messages by pointer, it needs a GVT/LBTS implementation to organise garbage collection. This inevitable delay explains the higher memory usage compared to adevs.\\
%TODO next sentence needs rewriting --done?
Optimistic's TimeWarp requires state/event saving, and its GVT algorithm is more complex (with a resultant higher latency) than the LBTS calculation in conservative. 
Additionally, the differences in LP virtual times are far larger compared to conservative time synchronization. All these factors explain the heavier memory usage. Devstone (flattened) is allocated in a chain. The leafs in the dependency graph will therefore do a lot of unnecessary simulation before having a revert, leading to quite severe memory pressure. Unlike conservative and sequential execution, memory usage in optimistic varies greatly depending on scheduling of kernel threads and drifting between kernels. 
\paragraph*{Interconnect}
In section 4.2 the parallel performance of this benchmark is further explained. It is of interest for sequential to contrast with the runtime penalty that dxex suffers w.r.t. adevs. 
\begin{table}[lhtb]
	\centering
	\caption{Interconnect w 10 t5e5, unit MiB, 2 kernels (if parallel)}
	\label{iconn_mem}
	\begin{tabular}{| l | l | l | l | l |}
		\hline
		Adevs & AdevsCon &DX &DXCon & DXOpt\\ \hline
		39 & 39 & 35 & 43 & 164 \\ \hline
	\end{tabular}
\end{table}

\paragraph*{Priority Network}
The priority network model is detailed in section 4.1.\\
Conservative's high memory usage is largely due to the garbage collection implementation. Kernel 0 is the initiator of the lbts calculation but since it holds only models independent of any other (server), it will finish very fast leaving the other kernel bereft of an updated lbts.
\begin{table}[lhtb]
	\centering
	\caption{Priority model n 128, m 16, p 10,  t2e8, unit MiB, 2 kernels}
	\label{pmod_mem}
	\begin{tabular}{| l | l | l |}
		\hline
		DX &DXCon &DXOpt\\ \hline
		35 & 450 & 651\\ \hline
	\end{tabular}
\end{table}

>>>>

%TODO say that you will now prove your claims about the aproach having nearly no impact
%TODO and potentially even claim that you are faster than adevs for some situations, due to the algorithms from PythonPDEVS and further features
%TODO briefly give some hardware info on the machine (CPU, OS, RAM), used software (version of adevs, GCC, ...), and about number of samples/variance
\subsection{Sequential Simulation}
%TODO we start with sequential simulation results, because this is still base case
%TODO compare CPU time with adevs for some benchmarks
%TODO and explain why there is such a difference
%TODO (spend some time here, as this is also important)
\subsubsection{Preamble}
The benchmark models are identical in functionality (between adevs/dxex versions). Dxex uses integer timestamps, adevs uses floating point. Tests show there is little difference performance wise (both types fit in a machine word)\\
Benchmarks ran on a i7-2670QM, 16GiB Ram, Arch Linux (4.2.5 kernel). Analysis of the benchmark results is done using call-graphs generated by perf, a selection of those call-graphs can be found in the code repository. \\
Note that a fixed timeadvance (e.g. devstone) is set at 100.
\subsubsection{Devstone}
The Devstone \cite{DEVStone} benchmark is highly hierarchical, using Direct connection the dxex kernels can exploit this whereas adevs needs to walk the coupled model's structure to pass events.\\
\begin{figure}[h]
	\includegraphics[width=.5\textwidth]{fig/fig1.eps}
	\label{fig1.eps}
\end{figure}

\subsubsection{PHold}
PHold \cite{PHOLD} is a parallel oriented benchmark, sequential runtime is measured only as a baseline. A model selects the destination of a message at runtime using a state-saved RNG, which takes up a non-trivial amount of runtime.

\subsubsection{Interconnect}
Interconnect \cite{van2013research} is a benchmark where all models broadcast, creating a complete graph in dependencies between models. As the model count increases, we see the expected quadratic increase in runtime for both adevs and dxex, but an increasing penalty for dxex. Profiling shows this is entirely due to the heap allocation of messages, which even though minimized by using memory pools remains significant.\\
\begin{figure}[h]
	\includegraphics[width=.5\textwidth]{fig/fig3.eps}
	\label{fig3.eps}
\end{figure}

\subsection{Parallel Simulation}
%TODO explain that we will now compare different synchronization protocols
%TODO compare different synchronization protocols with adevs' conservative, showing that we are faster (or at least, not that much slower) than adevs
\subsubsection{Devstone}
The flattened models are allocated to kernels by giving each kernel a distinct section of the chain, resulting in a low ratio of inter-kernel to intra-kernel messages. For optimistic, this can cause more reverts since the kernels will start to drift faster as the model-count increases. Furthermore, optimistic is quite sensitive to an increase in kernels, since the delay before a revert propagates increases. Of note as well is the warm-up time this benchmark requires, for $n=d\times w$ models, it takes $\textup{timeadvance}()*(n-1)$ transitions to activate the last model in the chain. For parallel this can be reduced to $\textup{timeadvance}()*(n\frac{kernels-1}{kernels} -1)$ before the last kernel becomes active.\\
\begin{figure}[ltbh]
	\includegraphics[width=.5\textwidth]{fig/fig2.eps}
	\label{fig2.eps}
\end{figure}
\subsubsection{PHold}
%Phold
In Phold allocation is specified in the benchmark itself. Each kernel manages a single node with a constant set of sub-nodes. The parameter R (here 10) determines the percentage of remote destination models.\\
The dynamic dependency graph is a very sparse version of the static dependency graph, penalizing conservative. Lookahead is $\epsilon$, so conservative spends most of its time crawling in steps of $\epsilon$. Since the dependency graph between kernels is a complete graph, this is not a simulation that scales in our implementation. For N kernels, each kernel has to query the null-time of N-1 kernels, resulting in O($N^2$) polling behaviour. In a non-cyclic simulation with a non-trivial lookahead (e.g. devstone), that choice however does pay off. Adevs's lesser performance is due in part to their lookahead management, which after profiling shows to spend a non-trivial amount of time in exception handling code. 
Optimistic suffers little from the above problems, however due to the high interconnectivity a cascading revert is still possible. With the percentage of remotes = 100 phold reflects interconnect in behaviour, which is why the R value is not dimensioned here. A revert is very expensive in PHold due to our usage of C++11's random nr generators. The cost of a revert is dominated by the recalculation of destination models, not in allocating/deallocating states/messages. Again, this could be significantly reduced using lazy-cancellation. Once a revert happens the drift between kernels increases fast, increasing the likelihood of more reverts. Despite all this, optimistic can for low R values quickly exploit uncertainty that slows down conservative in this benchmark.
\subsubsection{Interconnect}
In Interconnect
the set of atomic models form a complete graph (w.r.t. connections), each model broadcasts messages to the entire set. \\
Allocation cannot avoid cycles, the resulting dependency graph between kernels remains a complete graph. The runtime dependency graph is almost immediately equal to the static graph.
Conservative still faces the same issues as in PHold, with the key difference that for a fixed time advance lookahead is equal to the timespan between transitions. The scaling issue is identical as with PHold.\\
In this benchmark optimistic runs very quickly out of memory. With c kernels and N atomic models, a single revert undoing k transitions will lead to $(N-1)\times k$ messages that need to be recreated, plus $(N-1) \frac{c-1}{c} \times k$ anti-messages that need to be send.

\subsubsection{Priority network model}
%TODO show a single plot which varies a parameter that changes the ideal synchronization protocol
The priority benchmark is composed of a single server generating a stream of $0<=m<=n$ messages at fixed time intervals, interleaved with a probability p for a priority message, to n receivers. \\ This defaults lookahead for the receivers to $\epsilon$ but this time there is no scaling effect, nor are there cycles in the dependency graph. This model therefore highlights the basic strengths/weaknesses of both synchronization protocols. Receiving models are allocated on another kernel than the server, and have a internal transition so that they will not wait for the incoming messages. 
\input pqueue.tex \\
A key difference here with the other benchmarks is that a state (in the Receiver instances) is very cheap to copy/create. The kernel holding the server will never revert since it is a source in the dependency graph. Optimistic will therefore not suffer the same performance hit in recreating states as it does in PHold. The overhead in optimistic is entirely due to the factor m, which will quickly dominate in increasing buffers of received messages.
\begin{figure}[ltbh]
	\includegraphics[width=.5\textwidth]{fig/fig5.eps}
	\label{fig5.eps}
\end{figure}
Interestingly, the parameter p does not clearly benefit either synchronization protocol. While this removes any possibility for a lookahead, conservative can quickly bridge the timespan between fixed messages. Lightweight states prevent optimistic from performance loss due to reverts, only the overhead in storing messages becomes dominant.

%TODO (keep explaining as much as space permits, as this is the evaluation of the core contribution)
\subsection{Memory Usage}
%TODO compare memory usage with adevs for some benchmarks
\subsubsection{Platform and tools}
Both dxex and adevs use tcmalloc as memory allocator. Additionally, dxex uses memory pools to further reduce the frequency of expensive system calls (malloc/free/sbrk/mmap/...). Tcmalloc will only gradually release memory back to the OS, whereas our pools will not do so at all. If memory has been allocated once, it is from a performance point of view better to keep that memory in the pool. This is one reason why memory utilization is best measured by peak allocation. Profiling is done using Valgrind's massif tool. \cite{Nethercote:2007:VFH:1273442.1250746}
Platform used for memory profiling has an i5-3317U Intel cpu and 8GiB RAM with a page size of 4,096KiB, running Fedora 22 (kernel 4.2.6).\\
\subsubsection{Measure}
Adevs passes messages by value (in container by reference), we pass by pointer. The runtime effects of this choice are already demonstrated in the interconnect benchmark, so in this section, we measure memory usage in number of allocated pages combining text, stack and heap memory for the program profiled. For the OS and/or user this is the actual memory footprint of the application. It is important to note that, especially in the case of optimistic, not all this memory is always in use by the kernels. During simulation, the pools will generally not return deallocated memory to the OS, but keep it for later reuse.
\subsubsection{Results}
\paragraph*{Devstone}
\begin{table}[lhtb]
	\centering
	\caption{Devstone 40x40 t5e5, unit MiB, 4 kernels (if parallel)}
	\label{dtone_mem}
	\begin{tabular}{| l | l | l | l | l |}
		\hline
		Adevs & AdevsCon &DX &DXCon &DXOpt\\ \hline
		44 & 70 & 42 & 75 & 363  \\ \hline
	\end{tabular}
\end{table}
Since conservative passes messages by pointer, it needs a GVT/LBTS implementation to organise garbage collection. This inevitable delay explains the higher memory usage compared to adevs.\\
%TODO next sentence needs rewriting --done?
Optimistic's TimeWarp requires state/event saving, and its GVT algorithm is more complex (with a resultant higher latency) than the LBTS calculation in conservative. 
Additionally, the differences in LP virtual times are far larger compared to conservative time synchronization. All these factors explain the heavier memory usage. Devstone (flattened) is allocated in a chain. Leafs in the dependency graph will therefore do a lot of unnecessary simulation before having a revert, leading to severe memory pressure. Unlike conservative and sequential execution, memory usage in optimistic varies greatly depending on scheduling of kernel threads and drifting between kernels. 
\paragraph*{Interconnect}
In section 4.2 the parallel performance of this benchmark is further explained. Interconnect highlights dxex's lower performance due to message allocation overhead.
\begin{table}[lhtb]
	\centering
	\caption{Interconnect w 10 t5e5, unit MiB, 2 kernels (if parallel)}
	\label{iconn_mem}
	\begin{tabular}{| l | l | l | l | l |}
		\hline
		Adevs & AdevsCon &DX &DXCon & DXOpt\\ \hline
		39 & 39 & 35 & 43 & 164 \\ \hline
	\end{tabular}
\end{table}

\paragraph*{Priority Network}
The priority network model is detailed in section 4.1.\\
Conservative's high memory usage is largely due to the garbage collection implementation. Kernel 0 is the initiator of the lbts calculation but since it holds only models independent of any other (server), it will finish simulation very fast leaving the other kernel bereft of an updated lbts. The kernel holding the server will after simulation wait on the receiving kernel until it can prove all send messages can be deallocated.
\begin{table}[lhtb]
	\centering
	\caption{Priority model n 128, m 16, p 10,  t2e8, unit MiB, 2 kernels}
	\label{pmod_mem}
	\begin{tabular}{| l | l | l |}
		\hline
		DX &DXCon &DXOpt\\ \hline
		35 & 450 & 651\\ \hline
	\end{tabular}
\end{table}

### 2015-12-05 paper/6-conclusions.tex
Update lost profiles. Expand 4/6 a bit more + cite GVT futures.

%TODO conclude the paper, saying that our new simulation tool offers conservative and optimistic synchronization, resulting in good performance, which is even better than adevs
Both synchronization implementations offer good performance in differing simulations, in some simulations dxex outperforms adevs whereas in others we can still improve. 
The optimistic implementation uses an aggressive TimeWarp variant, this needs to be extended with lazy evaluation/cancellation to function in cyclic simulations. 
\subsection{Future work}
%TODO future work: more things that we might want to implement, and probably something about automatically switching between synchronization protocols at runtime (and reference my activity papers ;))
\subsubsection{Activity}
As shown in \cite{PythonPDEVS_ACTIMS} activity and allocation of models across kernels is a key aspect in achieving high performance in any parallel implementation. Allocating models so that there are no dependency cycles between their containing kernels is a first step, but not always possible. For optimistic one can use re-allocation to break (runtime dependency) cycles and/or perform load-balancing. If kernels are unevenly balanced they will begin to drift fast, causing increasingly more reverts. There is a limited framework implemented to track model activity for debugging purposes, this could be extended to enable the above strategies.
\subsubsection{Hybrid}
The optimistic implementation could use (null/eot/eit) from conservative to detect and/or reduce the cost of reverts without completely stalling on influencing kernels.
Conservative kernels could be extended with runtime information about influencing models, for example if one can guarantee a static dependency is not used for a fixed time-span, this can be removed for that period of (virtual) time. 
Ultimately the simulation could switch at runtime between protocols based on the information provided by activity tracking. This requires the above mentioned activity tracking framework, and could only be done if a GVT/LBTS time is agreed upon between the kernels. From the model point of view no changes are needed, although a non-trivial lookahead is obviously desired.

>>>>

%TODO conclude the paper, saying that our new simulation tool offers conservative and optimistic synchronization, resulting in good performance, which is even better than adevs
Sequentially dxex outperforms adevs in hierarchical simulations while yielding to adevs in broadcast-like simulations. 
Parallel the results are more nuanced. The sensitivity to broadcast simulations is carried over in the parallel implementation, resulting in performance loss for both synchronization protocols. Optimistic can still achieve good performance if the volume of inter-kernel messages is low. Both  are fast in non-cyclic simulations, with conservative better suited whenever state-saving becomes expensive and optimistic when lookahead is not available.\\
Finally, the current optimistic implementation is ill-suited for memory constrained systems, especially in simulations where the frequency of reverts is high.
\subsection{Future work}
%TODO future work: more things that we might want to implement, and probably something about automatically switching between synchronization protocols at runtime (and reference my activity papers ;))
\subsubsection{Memory}
Our optimistic implementation can benefit from a faster GVT algorithm such as FujiMoto's \cite{Fujimoto:1997:CGV:268403.268404} or more recently \cite{Bauer:2005:SND:1069810.1070159}. 
With access to the memory subsystem in place (pools), optimistic could be constrained using per kernel quota preventing the simulation from exhausting memory. While less sensitive to memory pressure, the same approach could benefit conservative.
\subsubsection{Activity}
As shown in \cite{PythonPDEVS_ACTIMS} activity and allocation of models across kernels is a key aspect in achieving high performance in any parallel implementation. Allocating models so that there are no dependency cycles between their containing kernels is a first step, but not always possible. For optimistic one can use re-allocation to break (runtime dependency) cycles and/or perform load-balancing. If kernels are unevenly balanced they will begin to drift fast, causing increasingly more reverts. There is a limited framework implemented to track model activity for debugging purposes, this could be extended to enable the above strategies.
\subsubsection{Hybrid}
The optimistic implementation could use (null/eot/eit) from conservative to detect and/or reduce the cost of reverts without completely stalling on influencing kernels.
Conservative kernels could be extended with runtime information about influencing models, for example if one can guarantee a static dependency is not used for a fixed time-span, this can be removed for that period of (virtual) time. 
Ultimately the simulation could switch at runtime between protocols based on the information provided by activity tracking. This requires the above mentioned framework, and could only be done at timepoints where a GVT/LBTS time is agreed upon between the kernels. From the model point of view no changes are needed, although a non-trivial lookahead is obviously desired.
